---
title: "motani_response"
output: pdf_document
date: "2023-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Response Motani et al 2022

R version 4.1.2 (2021-11-01) -- "Bird Hippie"

## Load packages

```{r library, message = FALSE, warning=FALSE}
library(ape)
library(geiger)
library(phytools)
library(OUwie)
library(reshape2)
library(ggplot2)
library(tidyr)
library(dplyr)
library(ggridges)

set.seed(2803)

```

## load data

```{r data}
dat_MR <- read.table("trait.txt",header=TRUE, sep = "\t")

MetRate <- dat_MR$Calculated.MRs..mL.O2...1.h...1.g.
names(MetRate) <- dat_MR$Taxon

phy <- read.nexus("consensus_Paleotree_final.tre")

phydata <- geiger::treedata(phy, MetRate, sort=TRUE)

MetRate_phy <- phydata$data
colnames(MetRate_phy) <- "MetRate"
```



## Original reconstruction

```{r originalrec}
fitBM_fastanc <-fastAnc(phydata$phy,MetRate_phy,vars=TRUE,CI=TRUE)
contmap_BM <- contMap(phydata$phy, MetRate_phy[,1], method = "user", 
                      anc.states = fitBM_fastanc$ace, plot = FALSE)
#change default colors
ramp_rate <- colorRampPalette(c("#02b2ce","#ffd004",  "#e52920"), bias=1.5) 
contmap_BM$cols[] <- ramp_rate(1001)

plot(contmap_BM, lwd = 2, fsize=0.5, outline=FALSE)
title(main="Original reconstruction - BM")
```


## How well fastAnc performs ASR ----

following:
<https://lukejharmon.github.io/ilhabela/instruction/2015/07/03/ancestral-states-1/>
(see also Revell, L. J., Harmon, L. J. (2022). Phylogenetic Comparative Methods in R. United Kingdom: Princeton University Press)

we simulate states in our phylogeny under BM

```{r correlation}
MetRate_BM <- fitContinuous(phydata$phy, MetRate_phy, model = "BM")
x.BM<-fastBM(phydata$phy,internal=TRUE, sig2= MetRate_BM$opt$sigsq,
             a= MetRate_BM$opt$z0)


## ancestral states
a.BM<-x.BM[as.character(1:phydata$phy$Nnode+Ntip(phydata$phy))]
## tip data
x.BM<-x.BM[phydata$phy$tip.label]

fit.test.BM <-fastAnc(phydata$phy,x.BM,vars=TRUE,CI=TRUE)

correl.BM <- cor(a.BM,fit.test.BM$ace)
```


```{r plotcor, echo=TRUE}
plot(a.BM,fit.test.BM$ace,xlab="true states",ylab="estimated states")
lines(range(c(x.BM,a.BM)),range(c(x.BM,a.BM)),lty="dashed",col="red") ## 1:1 line
title(main= paste("BM cor =", correl.BM))
```



They show high correlation, demonstrating that the function recovers credible ancestral states




## BAYESIAN ANALISIS ----

```{r bayesian}
set.seed(2803)
MetRate_phy.vec <- as.vector(MetRate_phy)
names(MetRate_phy.vec) <- rownames(MetRate_phy)
fit.bayes <- anc.Bayes(phydata$phy,MetRate_phy.vec, ngen = 99999)
plot(fit.bayes$mcmc$logLik)


#burning
fit.bayes.burned <- fit.bayes$mcmc[-c(1:200),]

bayes.res <- fit.bayes$mcmc[-c(1:200),-c(1,2,ncol(fit.bayes.burned))]
nodes.inference <- colMeans(bayes.res)
nodes.range <- lapply(bayes.res, range)
nodes.CI <- sapply(bayes.res, 
                   quantile, probs=c(0.025, 0.975))

nodes.inference.df <- data.frame(variable = names(nodes.inference), Bayesian=nodes.inference,
                                 CImin=nodes.CI[1,], CImax=nodes.CI[2,])

contmap_Bayes <- contMap(phydata$phy, MetRate_phy[,1], method = "user", 
                         anc.states = nodes.inference, plot=F)
contmap_Bayes$cols[] <- ramp_rate(1001)
plot(contmap_Bayes, lwd = 2, fsize=0.5, outline=F)

plot(fitBM_fastanc$ace,nodes.inference)
cor(fitBM_fastanc$ace,nodes.inference)
```

similar results than in our original reconstruction


## plot probability of every node to show that is not the same for all the confidence interval
```{r bayesplot}
bayes.res.long <- melt(bayes.res)

ggplot(data=bayes.res.long, aes(x = value)) + 
  facet_wrap(facets=vars(variable), nrow=6, scales="free")+
  geom_histogram()+
  geom_vline(dat=nodes.inference.df, aes(xintercept=Bayesian), col="red")+
  theme_classic()
```


## figure of Moteni et al but with probability of CI

```{r bayesplot2}
bayes.res.long2 <- melt(bayes.res[,c("73","74","77","86","88", "89","90","91","96","100","58")])


bayes.res.long2$Clade[bayes.res.long2$variable==73] <- 'Diapsida'
bayes.res.long2$Clade[bayes.res.long2$variable == 74] <- 'Choristodera'
bayes.res.long2$Clade[bayes.res.long2$variable == 77] <- 'Squamata'
bayes.res.long2$Clade[bayes.res.long2$variable == 86] <- 'Archosauria'
bayes.res.long2$Clade[bayes.res.long2$variable == 88] <- 'Ornithodira'
bayes.res.long2$Clade[bayes.res.long2$variable == 89] <- 'Pterosauria'
bayes.res.long2$Clade[bayes.res.long2$variable == 90] <- 'Dinosauria'
bayes.res.long2$Clade[bayes.res.long2$variable == 91] <- 'Ornithischia'
bayes.res.long2$Clade[bayes.res.long2$variable == 96] <- 'Saurischia'
bayes.res.long2$Clade[bayes.res.long2$variable == 100] <- 'Aves'
bayes.res.long2$Clade[bayes.res.long2$variable == 58] <- 'Mammalia'

bayes.res.long2$Clade <- factor(bayes.res.long2$Clade, level=c('Diapsida', 
                        'Choristodera', 'Squamata', 'Archosauria', 'Ornithodira', 
                        'Pterosauria', 'Dinosauria', 'Ornithischia', 'Saurischia', 
                        'Aves','Mammalia'))


node.means <- bayes.res.long2 %>% 
  group_by(Clade) %>% 
  summarise(meannode = mean(value))


ggplot(data=bayes.res.long2)+
  stat_density_ridges(aes(x=value, y=Clade, fill=stat(quantile)),
                      geom = "density_ridges_gradient", quantile_lines = FALSE, 
                      calc_ecdf = TRUE, scale=1,
                      quantiles = c(0.025, 0.975))+
  scale_fill_manual(name = "Prob.", values = c("white", "gray80", "white"))+
  stat_density_ridges(aes(x=value, y=Clade),fill="transparent",scale=1,
                      quantile_lines = TRUE,quantiles = c(0.5))+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90))+
  coord_flip()
```




# MODELS

## Now with an OU

```{r OUancml}
MetRate_phy.vec <- as.vector(MetRate_phy)
names(MetRate_phy.vec) <- rownames(MetRate_phy)
fitOU_ancml <-anc.ML(phydata$phy,MetRate_phy.vec, model = "OU")

contmap_OU <- contMap(phydata$phy, MetRate_phy[,1], method = "user", 
                      anc.states = fitOU_ancml$ace, plot = FALSE)
contmap_OU$cols[] <- ramp_rate(1001)

plot(contmap_OU, lwd = 2, fsize=0.5, outline=F)
title(main="FastAnc - OU")


plot(fitBM_fastanc$ace,fitOU_ancml$ace)
cor(fitBM_fastanc$ace,fitOU_ancml$ace)
```

We obtain very similar results than in our original reconstruction



Let's try with other packages, as anc.mL has not be thoroughly tested for OU models
# OUwie package ----

first we need to do a discrete character reconstruction

```{r discrete}
MetRate_dis <- dat_MR$Categorical.strategy
names(MetRate_dis) <- dat_MR$Taxon
phydata_dis <- geiger::treedata(phy, MetRate_dis, sort=TRUE)

MetRate_dis <- phydata_dis$data
colnames(MetRate_dis) <- "MetRate_dis"
MetRate_dis[MetRate_dis==0] <- 2

asr_dis_ml <- make.simmap(phydata_dis$phy, MetRate_dis[,1], model = "ER", nsim=1000) #we use an equal rates model assuming changing states is equally probable
asr_dis_db_ml <- describe.simmap(asr_dis_ml)

col_pd <- setNames(c("#e52920", "#02b2ce"), c("1", "2"))
plot(asr_dis_db_ml, fsize=0.5, ftype="i", colors = col_pd)
```

Now we fit models with OUwie package

```{r models, message=FALSE, warning=FALSE}
ouwie_phy <- phydata_dis$phy
ouwie_phy$node.label <- as.numeric(asr_dis_db_ml$ace[1:54,1]>0.5)
ouwie_phy$node.label[ouwie_phy$node.label==0] <- 2

data_ouwie <- data.frame(Genus_species= rownames(MetRate_phy), Reg=MetRate_dis[,1], 
                         X=MetRate_phy[,1])

MetRate_BM1 <- OUwie(ouwie_phy, data_ouwie, model = "BM1",root.station=FALSE, 
                     algorithm = "three.point", scaleHeight =FALSE)
MetRate_BMS <- OUwie(ouwie_phy, data_ouwie, model = "BMS", root.station=FALSE, 
                     algorithm = "three.point", scaleHeight = FALSE)
MetRate_OU1 <- OUwie(ouwie_phy, data_ouwie, model = "OU1", root.station=FALSE, 
                     scaleHeight = FALSE,algorithm = "three.point")
MetRate_OUM <- OUwie(ouwie_phy, data_ouwie, model = "OUM", root.station=FALSE, 
                     algorithm = "three.point", scaleHeight = FALSE)


aicw(c(MetRate_BM1$AICc,MetRate_BMS$AICc,MetRate_OU1$AICc,MetRate_OUM$AICc))
#OUM is the best fitted model
Ouwie_OUM_anc <- OUwie.anc(MetRate_OUM, knowledge=TRUE )

contmap_OUM_ouwie <- contMap(phydata$phy, MetRate_phy[,1], method = "user", 
                            anc.states = Ouwie_OUM_anc$NodeRecon, plot = F)
contmap_OUM_ouwie$cols[] <- ramp_rate(1001)

plot(contmap_OUM_ouwie, lwd = 2, fsize=0.5, outline=F)


plot(fitBM_fastanc$ace,Ouwie_OUM_anc$NodeRecon)
cor(fitBM_fastanc$ace,Ouwie_OUM_anc$NodeRecon)
```

This model fitted better, but it implies that species with different metabolic rates are subject to different regimes of natural selection, which is not the expectation. This complex models would require an underlying hypothesis. 
Running every available model and choosing the best fit is not always the best choice, as we need to consider their biological meaningfulness.




